Stage-Splitting a Modal Language

### 2014-10-14 paper/example.tex
worked the splitting example into the intro

One notable property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  Therefore, if quickselect were to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in \ref{fig:qssplit}),
and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).  
Equationally, we can represent this relationship as 
\[
\mathtt{qSelect(l,k)} \equiv \mathtt{qSelect2(qSelect1(l),k)}
\]

\subsection{Splitting of Quickselect}

Here, we will manually split apart the \texttt{qSelect} function in order to give some intuition about how the operation works.

We start with the basic \texttt{qSelect}:

@@

Again our goal is to split this into \texttt{qSelect1} and \texttt{qSelect2} such that \texttt{qStaged l i = qSelect2 (qSelect1 l) i}.  
The first step in this process is to assume this conclusion at the recursive sites:

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => qSelect2 (`1`qSelect1 left`2`) k
      | EQ => h
      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
\end{lstlisting}

We then pull the stage \bbone\ portion of each branch of the comparison up into a single tuple.

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
      case compare k n of
        LT => qSelect2 (pi2 p) k
      | EQ => pi1 p
      | GT => qSelect2 (pi3 p) (k-n-1)`
\end{lstlisting}

We continue to pull up the precomputation and include more in it:

@@

>>>>

One notable property of quickselect is that the partitioning of the list, 
even through its recursive calls, depends only on the values of list elements, and not on the value of $k$.
To illustrate this, we present colorized version of \texttt{qSelect}:

@@

Here, the input $k$ and all parts of the computation that depend on it are colored blue,
whereas all of the parts of the computation that do not depend on $k$ are colored red.
The \texttt{qSelect} variable itself, due to its status as a function with a mixed-color body, is left black.

To be precise, our coloring strategy respects data dependence, but not control dependence---the 
first arguments to the recursive calls to \texttt{qSelect} are colored red,
despite appearing underneath a blue case statement.
Because of this coloring strategy and the definition of data dependence, 
information can flow from red code to blue code but not vice versa.

Observe the that helper function \texttt{partition} is colored entirely red.
This is where much of the work of the quickselect algorithm takes place, and it doesn't even depend on $k$!

This suggests an optimization: we can split \texttt{qSelect} into two functions, 
one for all of the red code and another for all of the blue code. 
We'll call them \texttt{qSelect1} and \texttt{qSelect2}.
Because of the dependence properties stated above, 
\texttt{qSelect2} will need only depend on the results of \texttt{qSelect1} and the input index $k$,
whereas \texttt{qSelect1} will only depend on the input list $l$.
Of course, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
Equationally, we can represent this relationship as simply 
\[
\mathtt{qSelect~l~k} \equiv \mathtt{qSelect2~(qSelect1~l)~k}
\]

With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from \texttt{qSelect}.
Because the original function is recursive,
the first step in this process is to assume this conclusion at the call sites.
This is performed in place:
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => qSelect2 (`1`qSelect1 left`2`) k
      | EQ => h
      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
\end{lstlisting}
We then pull the stage \bbone\ portion of each branch of the comparison up into a single tuple.
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
      case compare k n of
        LT => qSelect2 (pi2 p) k
      | EQ => pi1 p
      | GT => qSelect2 (pi3 p) (k-n-1)`
\end{lstlisting}
This last step is necessary, but we can perform it only because we're willing to ignore the control dependence.
We continue to pull up the precomputation and include more in it:

@@

The final version is shown in figure...
Note that it's basically a tree algorithm...
